---
title: "Causal ML"
author: "Josia Schramm"
date: "2024-03-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
df_orig <- read.csv("data/data.csv")
library(tidyverse)
library(DoubleML)
library(mlr3)
library(mlr3learners)
library(ranger)
```

## 1.  define ATE and show its identification in a randomized controlled trial

The Average Treatment Effect (ATE) measures the mean difference in potential outcomes between the treatment and control groups. It is defined as:

$$
ATE = E[Y(1) - Y(0)]
$$

where $E$ stands for the expected value, $Y(1)$ represents the potential outcome if the unit received the treatment, and $Y(0)$ represents the potential outcome if the unit did not receive the treatment.

## 2. Extract Variables
```{r }

# summary(df_orig)

# baseline characteristcs, treatment variable, outcome variables we want (chosen from the paper):
# baseline:
# MOMAGE, MOMPAR, Delivered_Mode, Birth_weight, Treat_time,
# Cow_milk_60d, Honey, Infant_formula, Water_boil, BFEED Vitamin_drops, 

# treatment variables
# TX

# outcomes:
# DEATH, SEPSIS, SEPSISA, SEPSISB, SEPSISC, SEPSISDEATH, 
# DIARRHEA, ABSCESS, OMPHA


df <- df_orig %>% 
  dplyr::select(MOMAGE, MOMPAR, Delivered_Mode, Birth_weight, Treat_time, Cow_milk_60d, Honey, Infant_formula, Water_boil, BFEED, Vitamin_drops, TX, DEATH, SEPSIS, SEPSISB, SEPSISC, SEPSISDEATH, DIARRHEA, ABSCESS, OMPHA)

# sapply(df, table)
# sapply(df, class)

# Clean :
# 1. Delivered_Mode to regular birth/assisted birth
# 2. Treat_time to 2 or less, 3 or more
# 3. BFEED to <1 hr, > 1hr
# 4. change Cow_milk_60d, Honey, Infant_formula, Water_boil, Vitamin_drops, TX, DEATH, SEPSIS, SEPSISB, SEPSISC, SEPSISDEATH, DIARRHEA, ABSCESS, OMPHA

df <- df %>% 
  mutate(Delivered_Mode = ifelse(Delivered_Mode == "Assisted vagin"|Delivered_Mode == "C-section", "Assisted delivery", Delivered_Mode)) %>% 
  mutate(Treat_time = as.character(ifelse(Treat_time < 3, 0, 1))) %>% 
  mutate(Treat_time = ifelse(Treat_time == "0", "Treatment start < 2d after birth", "Treatment start > 2d after birth")) %>% 
  mutate(BFEED = ifelse(BFEED != "Immediately < 1 hour after", "< 1 hour after birth", BFEED)) %>% 
  mutate(BFEED = ifelse(BFEED == "Immediately < 1 hour after", "> 1 hour after birth", BFEED)) %>%
  mutate(TX_num = TX) %>% 
  mutate(SEPSISDEATH_num = SEPSISDEATH) %>% 
  mutate(across(c(Cow_milk_60d, Honey, Infant_formula, Water_boil, Vitamin_drops, TX, DEATH, SEPSIS, SEPSISB, SEPSISC, SEPSISDEATH, DIARRHEA, ABSCESS, OMPHA), as.factor)) %>% 
  drop_na()
  
# sapply(df, table)
# sapply(df, class)



```


## 3. descriptive statistics
```{r }
vars <- c("MOMAGE","MOMPAR","Delivered_Mode","Birth_weight","Treat_time",  
"Cow_milk_60d","Honey", "Infant_formula","Water_boil","BFEED"         
 ,"Vitamin_drops","DEATH", "SEPSIS","SEPSISB" , "SEPSISC" , "SEPSISDEATH"  ,  "DIARRHEA" ,      "ABSCESS", "OMPHA")


t1 <- tableone::CreateTableOne(vars = vars,
                               strata = "TX",
                               data = df,
                               smd = TRUE)
t1_print <- print(t1)


```
## 4. ATE 

```{r }
# ATE via mean comparison and univariable OLS to get CIs
ATE_mean <- mean(df[df$TX_num ==1,]$SEPSISDEATH_num)- mean(df[df$TX_num ==0,]$SEPSISDEATH_num)

OLS_for_CI <- lm(SEPSISDEATH_num ~ TX_num, data = df)
confint_1 <- confint(OLS_for_CI)
lower_CI_1 <- confint_1[2,1]
upper_CI_1 <- confint_1[2,2]

# ATQ via OLS
OLS_ATE_reg <- lm(SEPSISDEATH_num~TX_num+ MOMAGE+ MOMPAR+ Birth_weight+ BFEED, data = df)
ATE_OLS <- OLS_ATE_reg$coefficients[2]

confint_2 <- confint(OLS_ATE_reg)
lower_CI_2 <- confint_2[2,1]
upper_CI_2 <- confint_2[2,2]

# ATE via partial linear regression via DoubleML

df_prep <- df %>% 
  select(TX_num, SEPSISDEATH_num, MOMAGE, MOMPAR, Delivered_Mode, Birth_weight, Treat_time, Cow_milk_60d, Honey, Infant_formula, Water_boil, BFEED, Vitamin_drops)

learner = lrn("regr.ranger", num.trees = 100, mtry = 5, min.node.size = 2, max.depth = 5)
ml_l = learner$clone()
ml_m = learner$clone()

obj_dml_data = DoubleMLData$new(df_prep, y_col="SEPSISDEATH_num", d_cols="TX_num")
dml_plr_obj = DoubleMLPLR$new(obj_dml_data, ml_l, ml_m)
dml_plr_obj$fit()

PLR_ATE <- dml_plr_obj$coef
lower_CI_3 <- dml_plr_obj$coef - (1.96*dml_plr_obj$se)
upper_CI_3 <- dml_plr_obj$coef + (1.96*dml_plr_obj$se)

# ATE via AIPW

ml_g = lrn("regr.ranger", num.trees = 100, mtry = 5, min.node.size = 2, max.depth = 5)
ml_m = lrn("classif.ranger", num.trees = 100, mtry = 5, min.node.size = 2, max.depth = 5)
obj_dml_data = DoubleMLData$new(df_prep, y_col="SEPSISDEATH_num", d_cols="TX_num")
dml_irm_obj = DoubleMLIRM$new(obj_dml_data, ml_g, ml_m)
dml_irm_obj$fit()

AIPW_ATE <- dml_irm_obj$coef
lower_CI_4 <- dml_irm_obj$coef - (1.96*dml_irm_obj$se)
upper_CI_4 <- dml_irm_obj$coef + (1.96*dml_irm_obj$se)

```



## 5. Plot CIs for different methods
```{r }
ATE_data <- data.frame(
  Method = c("Mean Comparison", "OLS", "PLR", "AIPW"),
  Estimate = c(ATE_mean, ATE_OLS, PLR_ATE, AIPW_ATE),
  CI_lower = c(lower_CI_1, lower_CI_2, lower_CI_3, lower_CI_4),
  CI_upper = c(upper_CI_1, upper_CI_2, upper_CI_3, upper_CI_4)
)


ggplot(ATE_data, aes(x=Method, y=Estimate, ymin=CI_lower, ymax=CI_upper)) +
  geom_point() +
  geom_errorbar(width=0.2) +
  theme_minimal() +
  labs(title="Comparison of ATE Estimates with Confidence Intervals",
       x="Method",
       y="ATE Estimate")
```




## 6. Discuss results

The results make sense, since we have randomized the treatment, and we do not expect confounders to be present. This means that we can read the correct effects simply by doing a mean comparison, and do not expect a change when using more sophisticated methods.

## 7. Estimate CATE using causal_forest()
```{r }
df_causalf <- df_prep %>% 
  select(-SEPSISDEATH_num, -TX_num, -Treat_time, -BFEED, -Delivered_Mode) %>% 
  mutate(across(c(Cow_milk_60d, Honey, Infant_formula, Water_boil, Vitamin_drops),as.numeric))

sapply(df_prep, class)

tau.forest <- grf::causal_forest(df_causalf, df_prep$SEPSISDEATH_num, df_prep$TX_num)

CATE <- grf::average_treatment_effect(tau.forest, target.sample = "all")

```